---
title: "Assignment 5"
subtitle: "NHL Draft Analysis"
author: "Morjina Solaiman"
date: "2025-03-20"
categories: [Assignment, Dataviz]
code-fold: true
code-tools: true
format: html
---

## NHL Draft Analysis

### Load the dataset and libraries

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(readxl)
```

```{r}
#| message: false
#| warning: false
#Load the data
Individual.Skater <- read.csv("SkaterIndividualstats.csv")
OnIce.Skater <- read.csv("SkaterOnicestats.csv")
Goalie <- read.csv("Goalies.csv")
Individual.Skater.Rookie <- read.csv("RookieSkaterindividual.csv")
OnIce.Skater.Rookie <- read.csv("RookieSkaterOnIce.csv")
Rookie.Goalie <- read.csv("RookieGoalies.csv")
```

## VEZINA TROPHY

Presented to the goaltender “adjudged to be the best at this position.” NHL general managers vote on this award.

```{r}
#| message: false
#| warning: false
Scale.Goalies <- Goalie %>%
  filter(GP>25)

ggplot(Scale.Goalies, aes(x = reorder(Player, SV.), y = SV.))+
  geom_bar(stat = "identity", fill = "blue")+
  coord_flip()
ggplot(Scale.Goalies, aes(x = reorder(Player, -GAA), y = GAA))+
  geom_bar(stat = "identity", fill = "navyblue")+
  coord_flip()

```

## James Norris Memorial Trophy

Awarded to the defenseman who demonstrates “the greatest all-around ability” at the position.

```{r}
#| message: false
#| warning: false
Norris.OnIce <- OnIce.Skater%>%
  filter(Position == "D", GP > 30)%>%
  mutate(scale.CF. = scale(CF.),
         scale.xGF. = scale(xGF.))%>%
  select(Player, Team, GP, CF., xGF., scale.CF., scale.xGF. )
Norris.Indiv <- Individual.Skater%>%
  filter(Position == "D", GP > 30)%>%
  mutate(scale.Goals = scale(Goals),
         scale.Assists = scale(Total.Assists))%>%
  select(Player, Team, GP, Goals, Total.Assists, scale.Goals, scale.Assists )

Norris<- Norris.OnIce%>%
  left_join(Norris.Indiv, by = "Player")%>%
  mutate(scale.Norris = scale.CF. + scale.xGF. + scale.Goals + scale.Assists)

ggplot(Norris %>% filter(scale.Norris > 3), aes(x = scale.Goals, y = scale.Norris, label = Player)) +
  geom_point(aes(color = scale.Norris), size = 3) +
  geom_text(nudge_y = 0.05, size = 3) +  # Add player names to the scatter plot
  scale_color_gradient(low = "lightblue", high = "darkblue") +
  labs(
    x = "Scaled Goals",
    y = "Norris Score",
    title = "Top Players for the James Norris Trophy",
    subtitle = "Players with the highest Norris scores (2024-2025 season)") +
  theme_minimal()

```

## Hart Memorial Trophy

Awarded to the “player judged most valuable to his team.” This isn’t necessarily the best overall player, but rather the one who contributes most significantly to his team’s success.

```{r}
#| message: false
#| warning: false
Scale.OnIce <- OnIce.Skater %>% 
  filter(GP > 40) %>% 
  select(Player, Position, Team, CF., xGF.) %>%
  mutate(Hart = xGF. + CF.)

Top30.Players <- Scale.OnIce %>%
  arrange(desc(Hart)) %>%  # Sort players by Hart score in descending order
  head(30)

ggplot(Top30.Players, aes(x = reorder(Player, Hart), y = Hart, size = xGF., color = Position)) +
  geom_point() +  # Use points to represent players
  coord_flip() +  # Flip the coordinates to make the plot horizontal
  labs(x = "Player", y = "HART Score", size = "xGF Score", color = "Position") +
  ggtitle("Top 30 Players by HART Score and xGF Score") +
  theme_minimal()

```

## Calder Memorial Trophy

Given to the player “adjudged to be the most proficient in his first year of competition.” This is essentially the rookie of the year award.

```{r}
#| message: false
#| warning: false
Calder <- Individual.Skater.Rookie %>% 
  filter(GP > 35, Goals > 10 )

ggplot(Calder, aes(x = reorder(Player, Total.Points), y = Total.Points))+
  geom_bar(stat = "identity", fill = "skyblue")+
  coord_flip()
```

## Frank J. Selke Trophy

```{r}
#| message: false
#| warning: false
Selke <- OnIce.Skater %>% 
  filter(Position != "D", GP > 30, TOI > 1000 ) %>% 
  mutate(Selke. = CF+SF) %>% 
  filter(Selke. > 2000)

ggplot(Selke, aes(x = reorder(Player, Selke.), y = Selke.)) +
  geom_boxplot(
    fill = "lightgreen",  # Color for the box
    color = "darkgreen",  # Color for the border of the box
    outlier.color = "red",  # Color for outliers
    outlier.shape = 16,  # Shape for outliers
    outlier.size = 3  # Size of the outliers
  ) +
  coord_flip() +
  labs(
    x = "Player", 
    y = "Score", 
    title = "Best Players for the Selke Trophy (2024-2025 Season)"
  ) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(size = 8)
  )
```

## Lady Byng Memorial Trophy

Presented to the player who exhibits “the best type of sportsmanship and gentlemanly conduct combined with a high standard of playing ability.”

```{r}
#| message: false
#| warning: false
Lady <- Individual.Skater %>% 
  filter(TOI > 1000, Total.Points > 40, PIM < 20)

ggplot(Lady, aes(x = Total.Points, y = PIM, label = Player)) +
  geom_point(aes(color = Total.Points), size = 3) +
  geom_text(nudge_y = 1, size = 3) +
  labs(
    title = "Lady Byng Trophy Candidates",
    x = "Total Points",
    y = "Penalty Minutes (PIM)",
    color = "Total Points"
  ) +
  scale_color_gradient(low = "lightblue", high = "darkblue") +
  theme_minimal()
```