---
title: "Assignment 5"
subtitle: "NHL Draft Analysis"
author: "Morjina Solaiman"
date: "2025-03-03"
categories: [Assignment, Dataviz]
code-fold: true
code-tools: true
format: html
---

## NHL Draft Analysis

### Load the dataset and libraries

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(readxl)
NHLdraft <- read.csv("NHLdraft.csv")
NHLdraftstats <- read.csv("NHLdraftstats.csv")
NHLdictionary <- read_xlsx("NHLDictionary.xlsx")

```

Visualization of games played by drafted player in 2022

```{r}
#| message: false
#| warning: false
draftNHL <- NHLdraftstats %>% 
  select(name, round, overall, draftyear, position) %>% 
  group_by(name, round, overall, draftyear, position)

draft2022 <- NHLdraft %>% 
  filter(draftyear == 2022 & postdraft == 0)
ggplot(draft2022, aes(x = round, y = NHLgames, color = overall))+
  geom_point()+
  geom_jitter(width = 0.2, alpha = 0.7, height = 0.5)+
  labs(title = "NHL draft Analysis of 2022",
       x = "Hockey Games Round",
       y = "NHLgames")

```

Figure1 : This figure shows the number of NHL games played by players drafted in June 2022, categorized by draft round. Each point represents a player, and jittering has been applied to reduce overplotting. The visualization highlights how most drafted players do not immediately play in the NHL

## Expanded Scatterplot

```{r}
#| message: false
#| warning: false
draft2018 <- NHLdraftstats %>% 
  filter(draftyear == 2018 & postdraft < 6)
ggplot(draft2018, aes(x = round, y = NHLgames, color = postdraft))+
  geom_point()



```

```{r}
#| message: false
#| warning: false
drafttot2018 <- draft2018 %>% 
  group_by(playerId,round, overall, position, name) %>% 
  summarise(totgames = sum(NHLgames))
ggplot(drafttot2018, aes(x = round, y = totgames))+
  geom_point()

```

## Scatterplot with overall draft position

```{r}
#| message: false
#| warning: false
ggplot(drafttot2018, aes(x = overall, y = totgames))+
  geom_point()
```

## Scatterplot summary

```{r}
#| message: false
#| warning: false
ggplot(drafttot2018, aes(x = overall, y = totgames))+
  geom_point()+
  geom_smooth()

```

```{r}
#| message: false
#| warning: false
drafttot <- NHLdraft %>% 
  filter(postdraft < 8) %>% 
  group_by(playerId, round, overall, position, name) %>% 
  summarise(totgames = sum(NHLgames))
ggplot(drafttot, aes(x = overall, y = totgames))+
  geom_smooth(se = FALSE)

```

Average number of games played as a function

```{r}
#| message: false
#| warning: false
NHLdraftstats <- read.csv("NHLdraftstats.csv")
EL <- NHLdraftstats %>% 
  filter(name == "Elias Lindholm")

ggplot(NHLdraftstats %>% 
         filter(position == "Forward"),
       aes(x = postdraft, y = points))+
  geom_smooth(aes(color = as.factor(round)))+
  geom_point(data = EL, aes(x = postdraft, y = points), color = "red")

```

```{r}
#| message: false
#| warning: false
EL <-  NHLdraftstats %>%
  filter(name == "Elias Lindholm")
pick_31_range <-  NHLdraftstats %>% 
  filter()


```

## Barplot

```{r}
#| message: false
#| warning: false
drafttot2018 %>% filter(totgames > 15 & position == "Forward") %>%  ggplot(aes( y = reorder(name, totgames), x = totgames))+
  geom_col(color = "white", fill = "blue4")+
  labs(title = "Total games played by individual player in 2018",
       y = "Player Names", x = "Total Games")+
  theme(axis.text.x = element_text(angle = 90),
        legend.title = element_blank())
```

## Stacked Bar Chart

```{r}
#| message: false
#| warning: false
ggplot(draft2018, aes(x = postdraft, y = NHLgames, fill = as.factor(-round)))+
  geom_col(position = "stack")
```

```{r}
#| message: false
#| warning: false
NHLdraft1 <- NHLdraft %>% 
  mutate(descround = desc(as.factor(round)))

rounds <- c("darkred", "blue", "red", "darkblue", "pink", "violet", "lightblue")

ggplot(NHLdraft, aes(x = postdraft, y = NHLgames, fill = as.factor(-round)))+
  geom_col(position = "stack")+
  labs(title = "Pattern of play after drafting",
       x = "PostDraft",
       y = "Number of NHLgames")+
  theme(legend.position = "none")
         
```
